---
title: "Dataset: Accident and Emergency attendances"
author:
- '[Tom Jemmett][tj_email], [The Strategy Unit][su_web]'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataset: Accident and Emergency attendances}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

_Edited by Zoë Turner 28 August 2026_

This vignette explains how to use the `ae_attendances` dataset in R, and also details where it comes from and how it is generated.

The data is sourced from [NHS England Statistical Work Areas][nhs_swa] and is available under the [Open Government Licence v3.0][ogl_v3].

The data contains all reported A&E attendances for the period April 2016 through March 2019

The dataset contains:

- __period:__ date the month that this activity relates to, stored as a date (1st of each month) in the universal format of yyyy-mm-dd
- __org_code:__ factor of [ODS][ods] code for the organisation that this activity relates to
- __type:__ factor the [Department Type][ae_types] for this activity, either 1, 2, or other
- __attendances:__ number of the number of attendances for this department type at this organisation for this month
- __breaches:__ number of the number of attendances that breached the 4 hour target
- __admissions:__ number of the number of attendances that resulted in an admission to the hospital

## First, load the data and inspect it

```{r load data, message=TRUE, warning=TRUE}
library(scales)
library(ggrepel)
library(lubridate)
library(dplyr)
library(forcats)
library(tidyr)
library(NHSRdatasets)

ae_attendances <- NHSRdatasets::ae_attendances

ae_attendances |>
  mutate(
  # set the period column to show in Month-Year as a character format
    period = format(period, "%b-%y"),
  # set the numeric columns to have a comma at the 1000's place
    across(
      c(attendances, breaches, admissions),
      scales::comma
    )
  ) |>
  # show the first 10 rows
  head(10) |>
  # format for display
  kableExtra::kable()
  
```

We can calculate the 4 hours performance for England as a whole like so:

```{r england performance}
england_performance <- ae_attendances |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    dplyr::across(c(attendances, breaches), sum),
    .groups = "drop"
  ) |>
 dplyr::mutate(
    performance = 1 - breaches / attendances
  )

# format for display
england_performance |>
  dplyr::mutate(
    # same format options as above
    period = format(period, "%b-%y"),
    dplyr::across(c(attendances, breaches), scales::comma),
    # this time show the performance column as a percentage
    performance = scales::percent(performance)
  ) |>
  # show the first 10 rows and format as a table
  head(10) |>
  kableExtra::kable()

```

We can now plot the monthly performance

```{r england performance plot}
ggplot2::ggplot(england_performance, ggplot2::aes(period, performance)) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::labs(
    x = "Month of attendance",
    y = "% of attendances that met the 4 hour standard",
    title = "NHS England A&E 4 Hour Performance",
    caption = "Source: NHS England Statistical Work Areas (OGL v3.0)"
  )
```

We can clearly see the "Winter Pressures" where performance drops.

We can also inspect performance for the [different types of department][ae_types]:

```{r england performance by type}
ae_attendances |>
  dplyr::group_by(period, type) |>
  dplyr::summarise_if(is.numeric, sum) |>
  dplyr::mutate(performance = 1 - breaches / attendances) |>
  ggplot2::ggplot(ggplot2::aes(period, performance, colour = type)) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  # facet_wrap(vars(type), nrow = 1) +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(
    x = "Month of attendance",
    y = "% of attendances that met the 4 hour standard",
    title = "NHS England A&E 4 Hour Performance",
    subtitle = "By Department Type",
    caption = "Source: NHS England Statistical Work Areas (OGL v3.0)"
  )
```

From this it appears as if only the type 1 departments have the seasonal drops, type 2 and "other" departments remain pretty consistent.

## What are the best and worst trusts for performance?

We could create a similar table of data for performance by each individual trust, but it would be useful to only look at trusts that have a type 1 department as it appears from the chart above that these departments have the largest variation.

```{r performance_by_trust}
performance_by_trust <- ae_attendances |>
  dplyr::group_by(org_code, period) |>
  # make sure that this trust has a type 1 department
  dplyr::filter(any(type == 1)) |>
  dplyr::summarise(
    dplyr::across(
      c(attendances, breaches),
      ~ sum(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    performance = 1 - breaches / attendances
  )

# format for display
performance_by_trust |>
  dplyr::mutate(
    period = format(period, "%b-%y"),
    dplyr::across(c(attendances, breaches), scales::comma),
    performance = scales::percent(performance)
  ) |>
  head(10) |>
  kableExtra::kable()
```

From this table we can calculate the overall performance by each trust and then organise the trusts by their overall performance.

```{r performance_by_trust_ranking}
performance_by_trust_ranking <- performance_by_trust |>
  dplyr::summarise(performance = 1 - sum(breaches) / sum(attendances), .by = org_code) |>
  dplyr::arrange(performance) |>
  dplyr::pull(org_code) |>
  as.character()

print("Bottom 5")
head(performance_by_trust_ranking, 5)

print("Top 5")
tail(performance_by_trust_ranking, 5)
```

```{r performance_by_trust top 5 bottom 5 plot}
performance_by_trust |>
  dplyr::mutate(
    org_code = forcats::fct_relevel(
      org_code,
      performance_by_trust_ranking
    )
  ) |>
  dplyr::filter(org_code %in% c(
    head(performance_by_trust_ranking, 5),
    tail(performance_by_trust_ranking, 5)
  )) |>
  ggplot2::ggplot(aes(period, performance)) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::facet_wrap(vars(org_code), nrow = 2) +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(
    x = "Month of attendance",
    y = "% of attendances that met the 4 hour standard",
    title = "NHS England A&E 4 Hour Performance",
    subtitle = "Bottom 5/Top 5 over the whole 3 years",
    caption = "Source: NHS England Statistical Work Areas (OGL v3.0)"
  )
```

## Benchmarking

It is sometimes useful to see how an organisation stacks up against all of the other organisations. 
Below we create a chart where each organisation is shown as a point, ordered by performance from left (highest performance) to right (lowest) performance.

It's useful to indicate certain organisations on the chart, below I am showing the 3 organisations that are at the lower quartile, median and upper quartile, however you could change this to instead pick out specific organisations (using a reference table and `left_join` or hard coding with `case_when`).

```{r bencmarking plot}
ae_attendances |>
  dplyr::filter(period == last(period)) |>
  dplyr::group_by(org_code) |>
  dplyr::filter(any(type == 1)) |>
  dplyr::summarise_at(vars(attendances, breaches), sum) |>
  dplyr::mutate(
    performance = 1 - breaches / attendances,
    overall_performance = 1 - sum(breaches) / sum(attendances),
    org_code = forcats::fct_reorder(org_code, -performance)
  ) |>
  #
  dplyr::arrange(performance) |>
  # lets highlight the organsiations that are at the lower and upper quartile
  # and at the median. First "tile" the data into 4 groups, then we use the
  # lag function to check to see if the value changes between rows. We will get
  # NA for the first row, so replace this with FALSE
  dplyr::mutate(
    highlight = ntile(n = 4),
    highlight = tidyr::replace_na(highlight != lag(highlight), FALSE)
  ) |>
  ggplot2::ggplot(ggplot2::aes(org_code, performance)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = overall_performance)) +
  ggplot2::geom_point(ggplot2::aes(fill = highlight), show.legend = FALSE, pch = 21) +
  ggrepel::geom_text_repel(ggplot2::aes(label = ifelse(highlight, as.character(org_code), NA)),
    na.rm = TRUE
  ) +
  ggplot2::scale_fill_manual(values = c(
    "TRUE" = "black",
    "FALSE" = NA
  )) +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(),
    axis.ticks.y = ggplot2::element_line()
  )
```

[//]: <> (URL's / References --------------------------------------------------)
[tj_email]: mailto:thomas.jemmett@nhs.net
[su_web]:   https://www.strategyunitwm.nhs.uk/ 
[nhs_swa]:  https://www.england.nhs.uk/statistics/statistical-work-areas/ae-waiting-times-and-activity/
[ogl_v3]:   https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/
[ods]:      https://digital.nhs.uk/services/organisation-data-service
[ae_types]: https://web.archive.org/web/20200128111444/https://www.datadictionary.nhs.uk/data_dictionary/attributes/a/acc/accident_and_emergency_department_type_de.asp
