---
title: "Dataset: AphA CPD Survey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Dataset: AphA CPD Survey}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(NHSRdatasets)
library(tibble)
library(purrr)

apha_cpd_survey <- NHSRdatasets::apha_cpd_survey
```

This vignette details why the `apha_cpd_survey` dataset was created and how to load it with some details on how to view some of the data.
For code used to extract and clean the data can be found in `vignette("apha_cpd_survey")`

The dataset contains:

- __respondent_id:__ integer unique id for each respondent
- __form_started_dttm:__ date time
- __form_ended_dttm:__ date time
- __q00a_age_bracket_cat:__ character of age brackets 
- __q01a_gender_cat:__ character for Male, Female and Prefer not to say
- __q02a_max_education_level_cat:__ character 
- __q03a_analyst_years_n:__ character
- __q04a_role_description_cat:__ character freetext
- __q05a_afc_band_cat:__ character
- __q06a_current_org_years_cat:__ character
- __q07a_future_health_work_cat:__ character
- __q08a_apha_aware_ind:__ character
- __q09a_apha_registered_cat:__ character
- __q10a_other_org_member_txt:__ character
- __q11a_nhs_cpd_aware_cat:__ character
- __q12a_cpd_time_at_work_cat:__ character
- __q13a_cpd_outside_work_ind:__ character
- __q13b_cpd_outside_work_days_txt:__ character freetext
- __q14a_mgr_cpd_supportive_cat:__ character
- __q15a_mgr_cpd_more_support_ind:__ character
- __q16a_org_cpd_supportive_cat:__ character
- __q17a_org_cpd_budget_cat:__ character
- __q17b_org_cpd_budget_txt:__ character freetext
- __q18a_org_study_leave_ind:__ character
- __q18b_org_study_leave_days_txt:__ character freetext
- __q19a_mgr_cpd_discuss_freq_cat:__ character
- __q20a_cpd_opps_nhs_intranet_ind:__ character
- __q20b_cpd_opps_org_website_ind:__ character
- __q20c_cpd_opps_blogs_ind:__ character
- __q20d_cpd_opps_twitter_ind:__ character
- __q20e_cpd_opps_linkedin_ind:__ character
- __q20f_cpd_opps_word_of_mouth_ind:__ character
- __q20g_cpd_opps_apha_website_ind:__ character
- __q20h_cpd_opps_other_txt:__ character freetext
- __q21a_cpd_opps_sources_txt:__ character freetext
- __q22a_org_nhs_type_cat:__ character
- __q23a_org_analyst_team_size_cat:__ character
- __q24a_org_analytics_influence_cat:__ character

## Association of Healthcare Analysts Continuous Professional Development survey data

Data was provided with permission from Rony Arafin in 2024, the then president of AphA, to include this dataset as part of training. 

The data had been collected through a survey of analysts which was analysed and published by AphA on [their website](https://www.aphanalysts.org/ltnws/nhs-at-risk-of-losing-a-generation-of-data-analysts/).
This is a good example of real survey collected data as it includes some standardised text and some freetext with missing values and combinations of numbers and text in some fields.

There are also 38 columns of data with partial questions which are related, for example, Q20 has parts a, b, c, d, e, f, g and h.

This data has a lot of structure to it for text based analysis but is a good dataset for practising data cleaning.

## Labels

The data has labels as well as column headers which can be seen in the RStudio environment panel as `attr(*, "label")= chr`

To view the attribute/label for one column

```{r}
attributes(apha_cpd_survey$q00a_age_bracket_cat)
```

To view all labels with the headers

```{r}
df <- tibble(
  variable = names(apha_cpd_survey),
  variable_label = map_chr(
    apha_cpd_survey,
    ~ attr(.x, "label") %||% NA_character_
  ),
  value_labels = map(
    apha_cpd_survey,
    ~ attr(.x, "labels")
  )
)
```


To view just the data a bit more cleanly

```{r}
tibble::glimpse(apha_cpd_survey)
```


## Categories for columns

As there are so many columns finding the cateorgies or unique data is possible with using a function and a `purrr` loop:

```{r eval=FALSE}
# Simple function that uses base R unique() to return the unique data from a column
unique_data <- function(data, column) {
  unique(data$column)
}

# maps across all the columns and gives unique data
purrr::imap(
  apha_cpd_survey,
  ~ unique(.x)
)
```

As some columns have freetext or lots of unique data (as for date time columns) not all of this will be useful so restricting to only those that are known to have a few categories will be more helpful

```{r}
data <- apha_cpd_survey |>
  # unselect the columns with dttm in the name
  dplyr::select(!dplyr::ends_with("dttm")) |>
  # unselect respondent_id as that is a unique number
  dplyr::select(-respondent_id) |>
  # freetext columns
  dplyr::select(-c(
    q04a_role_description_cat,
    q13b_cpd_outside_work_days_txt,
    q17b_org_cpd_budget_txt,
    q18b_org_study_leave_days_txt,
    q20h_cpd_opps_other_txt,
    q21a_cpd_opps_sources_txt
  ))

# Rerun the purrr loop to see the data
purrr::imap(
  data,
  ~ unique(.x)
)
```

## Cleaning the data

This is good data to see how organisations can be listed in multiple ways if the collection is freetext.

For example:

```{r}
apha_cpd_survey |>
  dplyr::select(q22a_org_nhs_type_cat) |>
  dplyr::distinct() |>
  dplyr::filter(q22a_org_nhs_type_cat %in% c("Nhse", "NHS England", "NHS England ", "NHSE"))
```

all refer to NHS England and whilst some are spelled correctly there may be trailing white space.
