---
title: "Dataset: Covid19"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataset: Covid19}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(NHSRdatasets)

covid19 <- NHSRdatasets::covid19
```

This vignette details why the `covid19` dataset was created and how to load it.

The dataset contains:

- __date_reported:__ Date in the universal format yyyy-mm-dd
- __continent:__ Factor
- __countries_and_territories:__ factor
- __country_territory_code:__ factor
- __population_2019:__ integer
- __cases:__ integer
- __deaths:__ integer

## Covid19 data

This data was included in the package in August 2021.

The information was collected at the time from [European Centre for Disease Prevention and Control](https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide) and the page is still available but archived (as of August 2026). 
The data was made available under the open licence, compatible with the  CC BY 4.0 license, further details available at \href{https://www.ecdc.europa.eu/en/copyright}{ECDC}.

Data were collated and published up to 14th December 2020, and was tidied before submission (no script for that work is available).
