---
title: "cqc_directory dataset"
author:
- '[Martine Wauben](mailto:martine.wauben@dhsc.gov.uk)'
date: "04/10/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cqc_directory dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(tidyverse.quiet = TRUE)
```

This vignette explains how to use the `cqc_directory` dataset in R, and also
details where it comes from and how it is generated.

The data is sourced from [the Care Quality Commission](https://www.cqc.org.uk/about-us/transparency/using-cqc-data)
and its archive is available on [Google Drive](https://drive.google.com/drive/folders/1Y6V6r-q2l4lJYKuZL0DXw25VDUTH6L4N). 

The data contains care provider locations and ratings as published monthly in 2019. 

The dataset contains data on locations and providers:

- __location__: location ID, name, ODS code, telephone number, web address, sector, inspection directorate, region, local authority, CCG, street address, parliamentary constituency, and coordinates.
- __provider__: provider ID, name, company/charity registration, telephone number, web address, sector, inspection directorate, region, local authority, street address, and parliamentary constituency.
code for the organisation that this activity relates to
- __pub_date__: The date that particular datapoint was published

First let's load some packages and the dataset and show the first 10 rows of data.

```{r load data, message=TRUE, warning=TRUE}
library(dplyr)
library(knitr)
library(tidyr)
library(ggplot2)
library(lubridate)

library(NHSRdatasets)

data("cqc_directory")

# format for display
cqc_directory %>%
  # show the first 10 rows
  head(10) %>%
  # format as a table
  kable()
```

We can calculate the total number of each type of rating at each datapoint like this:

```{r ratings over time}
rat_overtime <- cqc_directory %>%
  group_by(pub_date, location_latest_overall_rating) %>%
  summarise(number = dplyr::n()) %>%
  filter(!is.na(location_latest_overall_rating)) %>%
  ungroup() %>%
  mutate(pub_date = as.Date(pub_date))

rat_overtime %>%
  pivot_wider(id_cols = location_latest_overall_rating,
              names_from = pub_date,
              values_from = number) %>%
  kable
```

We can now plot, for each month, the proportion of locations that received each rating:

```{r ratings plot}
ggplot(rat_overtime, aes(pub_date, number, fill = location_latest_overall_rating)) +
  geom_bar(stat = 'identity', position = 'fill')+
  theme_minimal()+
  labs(x = "Data publication date",
       y = "Proportion",
       fill = "CQC rating",
       title = "CQC ratings as published",
       caption = "Source: Care Quality Commission")
```

Although the raw numbers show increases in each rating, this is primarily driven by more and more locations becoming available: proportionally, performance has remained fairly stable at a national level.

We can further break this down by region to better understand what is happening:

```{r ratings over time by region}
region_ratings_overtime <- cqc_directory %>%
  group_by(pub_date, location_region, location_latest_overall_rating) %>%
  summarise(number = dplyr::n()) %>%
  filter(!is.na(location_latest_overall_rating) & !(location_region %in% c('Unspecified', '(pseudo) Wales'))) %>%
  ungroup() %>%
  mutate(pub_date = as.Date(pub_date))
ggplot(region_ratings_overtime, aes(pub_date, number, fill = location_latest_overall_rating)) +
  geom_bar(stat = 'identity', position = 'fill')+
  theme_minimal()+
  facet_wrap(~location_region)+
  labs(x = "Data publication date",
       y = "Proportion",
       fill = "CQC rating",
       title = "CQC ratings as published",
       caption = "Source: Care Quality Commission")
```

More interesting patterns emerge now: some regions have growing proportions of 'Good' or 'Outstanding' 
locations, whereas others are seeing increases in locations requiring improvement.

## How many new care homes are being opened?

With an ageing population, there is increasing demand for care home beds. Is supply keeping up?

```{r care home openings}
carehome_openings <- cqc_directory %>%
  mutate(location_hsca_start_date = as.Date(location_hsca_start_date)) %>%
  filter(care_home == 'Y' & location_hsca_start_date >= '2014-01-01' & 
           !(location_region %in% c('Unspecified', '(pseudo) Wales')) &
           pub_date == max(as.Date(pub_date))) %>%
  mutate(location_month = floor_date(location_hsca_start_date, unit = 'month')) %>%
  group_by(location_month) %>%
  summarise(number = n(),
            beds = sum(as.integer(care_homes_beds))) %>%
  mutate(location_year = as.factor(format(floor_date(location_month, unit = 'year'), '%Y')),
         month = format(location_month,"%m"))
ggplot(carehome_openings, aes(x = month, y = number, colour = location_year, group = location_year))+
  geom_line()+
  theme_minimal()
```

It is likely that the 2019 timeseries drops off due to a delay between provision starting and CQC reporting data.
However, overall the number of new care homes opening is not increasing! Perhaps each care home is getting bigger; 
how many new beds are becoming available?

```{r new beds added}
ggplot(carehome_openings, aes(x = month, y = beds, colour = location_year, group = location_year))+
  geom_line()+
  theme_minimal()
```

This is a bit more encouraging: in 2016 and 2017, there was a good uptick in new care home beds. However, 
2018 did not perform quite as well. Still, 2019 appears to be going in the right direction!

