---
title: "NHSRdatasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHSRdatasets}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

```{r setup}
library(NHSRdatasets)
```

NHSRdatasets is a package developed originally by NHS-R Community for use in its free R training and for example code in blogs or other material. 
It's doesn't contain any sensitive information and for some datasets it  follows the layout and structure of the data whilst for others it is a way of accessing the freely available data from elsewhere.
In these circumstances the data may have been transformed (changed).

These datasets are static and are not updated. 
As well as being available through R as a `function` the data is also available in the R data format `.rda` which are available from the [data folder on GitHub](https://github.com/nhs-r-community/NHSRdatasets/tree/main/data).

## Using the package

Once installed the data can be used in an R session by typing:

```{r}
NHSRdatasets::ae_attendances
```

or load the package first

```{r}
library(NHSRdatasets)

ae_attendances
```

To create an `object` so making the data available as a data frame:

```{r}
data <- ae_attendances
```

## Datasets

This package currently provides static datasets for:

- ae_attendances 
- apha_cpd_survey
- covid19
- LOS_model
- ons_mortality
- ons_uk_population_2023
- stranded_data
- synthetic_news_data

Each data set has documentation for the function which can be called by either putting the cursor in the name and selecting F1 or by running the R code:

```{r}
?ae_attendances()
```

It's also possible to write this without the brackets:

```{r}
?ae_attendances
```

If the package hasn't been loaded it's still possible to find the help files by using two question marks to look `globally`

```{r}
??ae_attendances
```

## Examples of use

[A search for the code](https://github.com/search?q=org%3Anhs-r-community+library%28NHSRdatasets%29&type=code) `library(NHSRdatasets)` on the NHS-R Community GitHub returns a number of examples, the search criteria being:

```
org:nhs-r-community library(NHSRdatasets)
```

It's used in other package examples like [NHSRplotthedots](https://github.com/nhs-r-community/NHSRplotthedots) and [FunnelPlotR](https://github.com/nhs-r-community/FunnelPlotR) and training like [Git](https://github.com/nhs-r-community/git_training) and [Intermediate R](https://github.com/nhs-r-community/intermediate_R_training).
