---
title: "Dataset: ONS Weekly Provisional Registered Deaths in England and Wales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Dataset: ONS Weekly Provisional Registered Deaths in England and Wales}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(NHSRdatasets)

ons_mortality <- NHSRdatasets::ons_mortality
```

This vignette details why the `ons_mortality` dataset was created and how to load it.

# Deaths registered weekly in England and Wales, provisional

Provisional counts of the number of deaths registered in England and Wales, by age, sex and region, from week commencing 8th January 2010 to 3rd April 2020.

This data has been made available through [Office of National Statistics under the Open Government Licence](http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/) but is made available in wide form across separate year's Excel spreadsheets.
These were brought together, tidied in a way they could be merged together and produced in long form using the code in the `vignette("create_ons_mortality")` and a blog can be found on the [NHS-R Community website](https://nhsrcommunity.com/blog/building-the-ons-mortality-dataset.html).

This data continues to be made available by ONS but is not added to this dataset. 
Corrections may have occurred too between this data snapshot and what is available through the ONS website.

The dataset contains:

- __category_1:__ character with categories that relate to grouped data like Persons, Males and Females but also for some other parts of the Excel dataset that aren't headers like "average of same week over 5 years"
- __category_2:__ character with categoris for grouped ages, age bands and regions as well as text from Excel spreadsheets like "v 2013 (IRIS)"
- __counts:__ numeric as counts of deaths recorded
- __date:__ date format as yyyy-mm-dd
- __week_no:__ integer relating to the week number starting at the first week of the year

## Exploring the data 

Looking at the unique values in `category_1` using base R which produces the data output as a data frame

```{r}
unique(ons_mortality$category_1)
```

or using dplyr which produces the data output as a tibble

```{r eval = FALSE}
ons_mortality |>
  dplyr::distinct(category_1)
```

there is a category `NA` which may not have information that can be used. 
When filtering for `NA` using `dplyr` we can combine it with the base R function `is.na()`

```{r}
ons_mortality |>
  dplyr::filter(is.na(category_1))
```

Incidentally to remove rows with `NA` an `!` is needed before the function in order to negate it

```{r eval=FALSE}
ons_mortality |>
  dplyr::filter(!is.na(category_1))
```

Or the function `filter_out()` can be used which can make this easier to read as it may be easy to overlook a single character `!`

```{r eval=FALSE}
ons_mortality |>
  dplyr::filter_out(is.na(category_1))
```

There are 52 rows with `category_1` that is `NA` and to view them all that's possible in the console by extending the tibble view with

```{r}
library(magrittr)

ons_mortality |>
  dplyr::filter(is.na(category_1)) %>%
  print(n = nrow(.))
```

Notice that this code has used the pipe `%>%` which comes from the package `magrittr` which is also in `tidyverse`. 
This pipe, rather than the base R |> allows the data to pass to the next level which is seen as `.`.

It's not clear from this search what the counts 1779 and 1925 relate to (other than being from January 2014) and this can only be understood by going back to the original spreadsheets and data manipulation which can be found in the `vignette("create_ons_mortality")`.

## Small charts exploration

A line chart to see all the Regions using `ggplot2`

```{r}
by_region <- ons_mortality |>
  dplyr::filter(category_1 == "Region") |>
  ggplot2::ggplot(ggplot2::aes(date, counts, colour = category_2)) +
  ggplot2::geom_line() +
  ggplot2::geom_point()

by_region
```

In the view of small charts using the facet grid to show each area separately

```{r}
by_region + ggplot2::facet_wrap(~category_2)
```
