---
title: "Dataset: ONS UK population 2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataset: ONS UK population 2023}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(NHSRdatasets)

ons_uk_population_2023 <- NHSRdatasets::ons_uk_population_2023
```

This vignette details why the `ons_uk_population_2023` dataset was created and how to load it.

# ONS UK Population 2023

This dataset was added in October 2024 as part of Hacktoberfest.
The original data was sourced from the [ONS mid year estimates of population for England and Wales]( https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland) and was tidied using the R script in `data-raw\ons_uk_population_2023.R`.


The dataset contains:

- __sex:__ character for females and males
- __Code:__ character code for areas which corresponds to the Name column
- __Name:__ character for areas in the UK and in this dataset has been restricted to Countries and grouped Countries (United Kingdom and Great Britain)
- __Geography:__ character detail on area and in this dataset is just Country
- __age:__ character with age in years from 0 to 90+
- __count:__ number referring to the count of population 

